---
output:
  md_document:
    variant: markdown_github
---

# Purpose

This folder contains my solutions to the Data Science for Economics and Finance 
Practical Exam. Each question is contained in its own subfolder, with functions 
stored in the respective `code/` folders and data stored in the respective 
`data/` folders (gitignored).

---

# Question 1: Coffee Hub

## Approach

A coffee entrepreneur wants to know which origin, roast strength, and supplier to stock for a new shop, and how that lines up with what Stellenbosch students say they like in their coffee.

I approached this by building one function per plot, namely origin vs cost, roast strength, top suppliers, and a keyword-match check against the student survey wordcloud. I cleaned the data first (checked for missing values and tidied a few wide-format columns into long form), then built each plot to answer one specific part of the brief rather than trying to cram everything into a single chart.

```{r q1 setup}
library(tidyverse)

list.files('Question1/code/', full.names = T, recursive = T) %>%
    .[grepl('.R', .)] %>%
    as.list() %>%
    walk(~source(.))

Coffee <- silent_read(path = "Question1/data/Coffee/Coffee.csv")
```

### Data prep

Checked for missing values and a few messy duplicate names (e.g. roaster names with curly vs straight apostrophes) before tidying `origin_1`/`origin_2` and `desc_1`/`desc_2`/`desc_3` from wide to long format, since each is really one variable spread across multiple columns.

```{r}

# Check for missing values/NA's
Coffee %>% summarise(across(everything(), ~sum(is.na(.))))

Coffee_long <- Coffee %>%
    gather(origin_rank, origin, origin_1, origin_2) %>%
    gather(desc_rank, desc, desc_1, desc_2, desc_3) %>%
    filter(!is.na(desc))  # two coffees are missing a third review so those rows are dropped

# Chiriquí was splitting into 2 rows under different spellings, messing up the "best rated" label - fixed here so it applies to all the origin plots
Coffee_long <- Coffee_long %>%
    mutate(origin = gsub(" Growing Region| Growing District| District| Department| County| Province| Zone| Region", "", origin)) %>%
    mutate(origin = gsub("í", "i", origin))
```

### Origin vs cost

I plotted cost against rating to see if price actually tracks quality — bubble size shows review count, so a region's position isn't trusted equally regardless of how many reviews back it up.

Northern Sumatra is both the highest-rated origin and inexpensive, while Cobán shows a well-rated coffee can be found for under $4 — high rating doesn't require a high price here.

```{r}
g <- origin_cost_scatter(Coffee_long)
g
```

### Roast strength

A boxplot with individual reviews jittered on top shows both the typical rating per roast and how much it varies. This is useful since a roast could have a good median but huge spread.

Light and Medium-Light roasts consistently outscore Medium, Medium-Dark, and especially Dark roasts. 

```{r}
g <- roast_plot(Coffee)
g
```

### Top suppliers

Ranked suppliers by rating directly, since that's the simplest way to answer "who should she buy from". I filtered to a review-count minimum so high ratings (with a low number of them) don't dominate, and a cost ceiling so the list stays realistic for a student-shop having to compete with places like MyVrew which sell relatively cheap coffee.

Top-rated suppliers under $20/100g with at least 5 reviews each.

```{r}
g <- supplier_plot(Coffee)
g
```

### Matching student preferences

The brief gave a wordcloud of words Stellenbosch students used to describe coffee they liked. I picked the words that stood out most and checked which origins' reviews mention them the most.

Reviews from Yemen, Tarrazu, and Boquete mention student-preferred flavour words (chocolate, aroma, savory, toned, sweetly, tart) most often.

```{r}
g <- keyword_plot(Coffee_long)
g
```

---

# Question 2: Baby Names

## Approach

Brief description of approach.

## Figure 1


Interpretation.


## Figure 2


Interpretation.

---

# Question 3: Loans and Credit

## Approach

Brief description of approach.

---

# Question 4: Netflix

## Approach

Brief description. Note which aspects of the data were chosen to explore and why.

---


