---
output:
  md_document:
    variant: markdown_github
---

# Purpose

This folder contains my solutions to the Data Science for Economics and Finance Practical Exam. Each question is contained in its own subfolder, with functions stored in the respective `code/` folders and data stored in the respective `data/` folders (gitignored).

---

# Question 1: Coffee Hub

## Approach

A coffee entrepreneur wants to know which origin, roast strength, and supplier to stock for a new shop, and how that lines up with what Stellenbosch students say they like in their coffee.

I approached this by building one function per plot, namely origin vs cost, roast strength, top suppliers, and a keyword-match check against the student survey wordcloud. I cleaned the data first (checked for missing values and tidied a few wide-format columns into long form), then built each plot to answer one specific part of the brief rather than trying to cram everything into a single chart.

```{r q1 setup}
library(tidyverse)

list.files('Question1/code/', full.names = T, recursive = T) %>%
    .[grepl('.R', .)] %>%
    as.list() %>%
    walk(~source(.))

Coffee <- silent_read(path = "Question1/data/Coffee/Coffee.csv")
```

### Data prep

Checked for missing values and a few messy duplicate names (e.g. roaster names with curly vs straight apostrophes) before tidying `origin_1`/`origin_2` and `desc_1`/`desc_2`/`desc_3` from wide to long format, since each is really one variable spread across multiple columns.

```{r}

# Check for missing values/NA's
Coffee %>% summarise(across(everything(), ~sum(is.na(.))))

Coffee_long <- Coffee %>%
    gather(origin_rank, origin, origin_1, origin_2) %>%
    gather(desc_rank, desc, desc_1, desc_2, desc_3) %>%
    filter(!is.na(desc))  # two coffees are missing a third review so those rows are dropped

# Chiriquí was splitting into 2 rows under different spellings, messing up the "best rated" label - fixed here so it applies to all the origin plots
Coffee_long <- Coffee_long %>%
    mutate(origin = gsub(" Growing Region| Growing District| District| Department| County| Province| Zone| Region", "", origin)) %>%
    mutate(origin = gsub("í", "i", origin))
```

### Origin vs cost

I plotted cost against rating to see if price actually tracks quality — bubble size shows review count, so a region's position isn't trusted equally regardless of how many reviews back it up.

Northern Sumatra is both the highest-rated origin and inexpensive, while Cobán shows a well-rated coffee can be found for under $4 — high rating doesn't require a high price here.

```{r}
g <- origin_cost_scatter(Coffee_long)
g
```

### Roast strength

A boxplot with individual reviews jittered on top shows both the typical rating per roast and how much it varies. This is useful since a roast could have a good median but huge spread.

Light and Medium-Light roasts consistently outscore Medium, Medium-Dark, and especially Dark roasts. 

```{r}
g <- roast_plot(Coffee)
g
```

### Top suppliers

Ranked suppliers by rating directly, since that's the simplest way to answer "who should she buy from". I filtered to a review-count minimum so high ratings (with a low number of them) don't dominate, and a cost ceiling so the list stays realistic for a student-shop having to compete with places like MyVrew which sell relatively cheap coffee.

Top-rated suppliers under $20/100g with at least 5 reviews each.

```{r}
g <- supplier_plot(Coffee)
g
```

### Matching student preferences

The brief gave a wordcloud of words Stellenbosch students used to describe coffee they liked. I picked the words that stood out most and checked which origins' reviews mention them the most.

Reviews from Yemen, Tarrazu, and Boquete mention student-preferred flavour words (chocolate, aroma, savory, toned, sweetly, tart) most often.

```{r}
g <- keyword_plot(Coffee_long)
g
```

---

# Question 2: Baby Names

## Approach

This question asked for the analysis of US baby naming trends (1910-2014) for a toy design agency.The agency propositioned this in order to understand what drives naming trends and how long they persist, to inform toy character naming decisions.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
list.files('Question2/code/', full.names = T, recursive = T) %>%
    .[grepl('.R', .)] %>%
    as.list() %>%
    walk(~source(.))

BabyNames <- read_rds("Question2/data/US_Baby_names/Baby_Names_By_US_State.rds")
Billboard <- read_rds("Question2/data/US_Baby_names/charts.rds")
HBO_titles <- read_rds("Question2/data/US_Baby_names/HBO_titles.rds")
HBO_credits <- read_rds("Question2/data/US_Baby_names/HBO_credits.rds")
```

```{r inspect, include=FALSE}
#Have a look at each dataset to see what I am working with
glimpse(BabyNames)
glimpse(Billboard)
glimpse(HBO_titles)
glimpse(HBO_credits)

# Check for NA values
BabyNames %>% summarise(across(everything(), ~sum(is.na(.))))
```


# Naming Persistence Over Time

`calc_rank_persistence()` takes the top 25 names for a given year and sex, then looks up each of those names in the full national ranking three years later however far down they may have fallen. This is better than comparing two top-25 lists directly (which was my initial appraoch), because names that dropped out of the top 25 still get penalised rather than being silently excluded. `persistence_plot()` then plots the resulting Spearman correlation for every year from 1910 to 2011, separately for boys and girls.

```{r persistence-calc and plot}
Years <- BabyNames$Year %>% unique() %>% sort()
Years <- Years[Years <= max(Years) - 3]

PersistenceResult <-
    expand_grid(Year = Years, Gender = c("M", "F")) %>%
    mutate(Corr = map2_dbl(Year, Gender, ~calc_rank_persistence(BabyNames, .x, .y)))

g <- persistence_plot(PersistenceResult)
g
```

- Persistence dropped and became more volatile after 1990 (for both sexes)
- Girls' names persistence started weakening earlier (mid-1950s/60s, not 1990)
- Boys' names persistence stayed stable until about 1990, then matched the same pattern
- Persistence is mostly weaker and more volatile since 1990 than in the early decades (for both sexes)

# Naming Surges and Cultural Events

The brief's tip was to look for year-on-year surges and dig into what might explain them, with Whitney in the 1980s given as the model example. So I built `find_name_surges()` to find the biggest single-year jumps in national popularity.

```{r find-surges}
NameSurges <- find_name_surges(BabyNames, top_n = 15)
NameSurges
```

- Several biggest male surges (Robert, John, James, Richard, Michael) all
  happen in 1946, suggesting a shared cause rather than separate ones
    - Likely the post-WWII baby boom, which raised total births that year
- Female surges (Linda, Shirley, Ashley) are more spread out, suggesting
  separate, name-specific causes
    - Shirley likely tracks child star Shirley Temple's rise in the 1930s
    - Linda and Ashley probably follow similar celebrity-driven trends,
      though not individually verified

## More Recent Rising Names

To zoom into the most recent decade of data, I plotted the names with the biggest average rise in popularity between 2005 and 2014, separately for girls and boys, using my existing `plot_rising_names()` function. Rather than just listing the names with a single surge year, this shows their actual year-by-year counts as lines so it's possible to see each name rose, not just that it rose.

# More Recent Rising Names
```{r rising names per gender plots, echo=FALSE}
g <- plot_rising_names(BabyNames, gender_input = "F", from_year = 2005, to_year = 2014, top_n = 5)
g

g <- plot_rising_names(BabyNames, gender_input = "M", from_year = 2005, to_year = 2014, top_n = 5)
g

```

- 2005 to 2014's fastest-rising names (Sophia, Harper, Liam, Mason, etc.) show       steady multi-year growth rather than a single sharp spike
- Daleyza stands out as a name with zero history before 2012, then immediate growth   (a different pattern from names simply gaining popularity over time)
  - Interestingly, a quick google search attributes this rise in the name to the Mexican-American singer (Larry Hernandez) naming his daughter in 2010

One name in this list genuinely surprised me was **Daleyza**. I'd never even heard this name before, and it doesn't follow the same pattern as most of the other surges on the list because there's basically zero history before it suddenly shows up, rather than slowly building up like a normal "rising" name. That jump from nowhere is what had me running to google and it turns out Mexican-American singer Larry Hernandez named his daughter Daleyza in 2010, which seems to line up with the name appearing out of nowhere shortly after. He also had a reality show called LarryMania - the more you know...

This prompted me to check it out:

```{r}
Daleyza_by_state <- BabyNames %>% filter(Name == "Daleyza") %>%
    group_by(State) %>% summarise(Total = sum(Count)) %>% arrange(desc(Total))

Daleyza_by_state
```

And suprise, suprise! Daleyza's spike is heavily concentrated in California and Texas, with Arizona, Illinois, and Colorado following behind them. This matches the Larry Hernandez explanation because many Mexican-Americans live in thse states (but they are also just some of the most densely populated states in the U.S.)

## Conclusions
- Persistence has become less predictable since the 1960s, not specifically since 1990 as suspected by the agency. 
    - Both boys' and girls' rankings show clear volatility decades before 1990, so the 1990s aren't a clean turning point, but the bigger instinct (that recent name trends are less stable than older ones) holds for the second half of the dataset.
- A practical implication for toy naming is that a name sitting in the top 25 today is a weaker guarantee of future popularity than it would have been pre-1960s so betting heavily on one trending name for a toy line carries more risk now than in earlier decades.
- Large naming surges have two driving forces, and they require different responses by the agency:
    - Population-wide events (e.g. the 1946 male-name cluster, likely tied to the post-WWII baby boom) lift already-popular names rather than creating new fashionable ones - not useful for spotting a "next big name," since nothing new is actually trending
    - Culturally-driven, name-specific spikes (Shirley Temple, Whitney Houston and also likely Daleyza/Larry Hernandez) are the ones actually worth watching. 
    - These show a single celebrity, character, or cultural moment can cause a name to gain popularity fast. 
        - This is clearly the the driver that the agency should pay attention to for toy naming purposes
- The agency should also watch out for zero-history to sudden spike names like Daleyza as this could hint to a name suddenly gaining popularity. 
- Overall, in recent years names have not been as persistent and so toy companies' should maybe shy away from relying too much on such data.
- Caveat: this analysis focused on the baby-names data itself; a deeper dive cross-referencing HBO character names and Billboard artists/songs against naming spikes (as outlined in the brief) was not completed in this ietration due to time constraints


---

# Question 3: Loans and Credit

## Approach

This question asked for an analysis of US household loan default trends using anonymised Lending Club peer-to-peer lending data. The Director of the Credit Institute in Texas wanted to understand what drives defaults and whether four beliefs of the Institute hold up against the data.

```{r setup-data}
list.files('Question3/code/', full.names = T, recursive = T) %>%
    .[grepl('.R', .)] %>%
    as.list() %>%
    walk(~source(.))

# Load the data
Loan_credit <- read_rds("data/Loan_Cred/loan_data.rds")

# Look at the data to assess the necessary steps for cleaning
dim(Loan_credit)

glimpse(Loan_credit)

summary(Loan_credit)

# Apply data cleaning function
Loan_credit <- clean_loans(Loan_credit)
```

The data has 1 million rows and 145 columns, so the first thing I did was reduce it to only the columns actually needed for analysis. I then built a `clean_loans()` function to create the outcome variables, convert text columns to numeric, and Winsorise relevant (ie. variables with crazy values as seen from the summary function) continuous variables at the 1st and 99th percentiles. Since there is no age variable in the dataset, I used credit history length derived from `earliest_cr_line` as a proxy since I thought that a longer history likely means an older borrower.

# Key Risk Drivers

The dataset contains many variables that could serve as default risk drivers (revolving credit utilisation (`revol_util`), number of public records (`pub_rec`), debt-to-incomeratio (`dti`), income verification status (`verification_status`), and annual income (`annual_inc`) are all standard credit risk indicators a lender would most likely consider. However, I focused on credit grade and loan purpose because they are the most immediately interpretable at a glance as a director looking at a bar chart of default rates by grade or by loan purpose can draw actionable conclusions without needing to understand the underlying variable in detail. 

```{r q3-grade, echo=FALSE}
defaults_by_grade_plot(Loan_credit)
```

The plot shows a clear strong relationship between credit grade and default rate (as expected). 

```{r q3-purpose, echo=FALSE}
defaults_by_purpose_plot(Loan_credit)
```

Car and home improvement loans carry the lowest default rates at around, while small business and wedding loans sit at the top at. Thus, lenders should apply greater scrutiny to higher-risk purpose categories when assessing borrower credibility. This plot provides evidence that loan purpose is a a good credit risk signal. 

# Looking at the Beliefs held by the Institute

The Director wishes to challenge four core beliefs held by the Institute so I tested each one using the cleaned data, building a separate plot for each belief.

The first belief was that homeowners and people employed for more than 10 years default far less on short-term loans. To test this I built a line plot showing default rates across home ownership categories, with separate lines for borrowers employed 10+ years vs under 10 years, filtering to short-term loans only since the belief is specifically about those. The data contains only two loan terms (36 and 60 months) so 36 months is used as the short-term definition.

```{r homeowner-employment, echo=FALSE}
homeowner_employment_plot(Loan_credit)
```

The plot somewhat confirms this belief since borrowers employed for 10 or more years show slightly lower default rates (lower by $1%$). This difference is really small so this belief is only partly supported by the data. On the other hand, homeowners (those with over and under 10 years of employment) have higher default rates than borrowers.

The second belief was that US states differ significantly in their culture of defaulting. To test this I plotted default rates for every state ordered from lowest to highest, with Texas highlighted separately so it is easy to see where it sits relative to the rest.

```{r default_by_state, echo=FALSE}
default_by_state_plot(Loan_credit)
```

This belief is more closely supported by the data because default rates range from around $14%$ to about $27%/28%$. This means that US states do in fact differ significantly in default culture. Texas sits close to the average when looking at the bigger picture. 

The third belief was that credit grades predict default well specifically for younger individuals. Since there is no age variable in the data I used credit history length as a proxy since shorter history likely means a younger borrower. I then plotted default rates by grade for each history group using `facet_wrap`, which creates one panel per group so the grade pattern can be compared side by side.

```{r grade-age, echo=FALSE}
grade_age_plot(Loan_credit)
```

The credit grading system has the same relationship as before for both younger (shorter credit history) and older (longer credit history) individuals. This relationship is that default rates increase with lower credit grades. The differences between older and younger individuals default rates' is very small. For higher grades, younger individuals have a slightly higher default rating while for higher grades it is the opposite. Thus, there is no evidence from the data to support the belief that grades work better or worse for younger borrowers specifically.

The fourth belief was that interest rates are determined by age, occupation 
and credit scores. Due to time constraints, I focused on credit grade as a proxy for credit score and plotted the median interest rate per grade. I will come back to this if I have the time - possible to plot age (as proxied by credit history) against median interest rates. 

```{r int-rate, echo=FALSE}
interest_rate_plot(Loan_credit)
```

I wanted to use the credit history length as a proxy for age here again, but I got stuck and am cautious of the time. I will return to this if I have time left. 

From this graph it is obvious that credit rating is an obvious determinant of interest rates. From the basic understanding of interest rates and credit ratings, this correlation makes sense, because a lower credit rating means there is a higher risk of default. Thus, the risk is compensated for by a higher interest rate (compared to higher credit ratings). The dataset does not have any specific age or occupation data so it would be difficult to draw conculsions about those factors. 


## A note on knitting

I unfortunately spent too long tweaking the Texevier YAML template for Question 3 and broke it (sorry) in the process. I get a `\pandocbounded` LaTeX error which I will torubleshoot if I have time. The analysis and all functions run correctly in RStudio, the PDF just would not compile in time. The full write-up lives in `Question3/Question3.Rmd` if needed.
---

# Question 4: Netflix

## Approach

Brief description. Note which aspects of the data were chosen to explore and why.

---


