---
output:
  md_document:
    variant: markdown_github
---

# Purpose

This folder contains my solutions to the Data Science for Economics and Finance Practical Exam. Each question is contained in its own subfolder with it's own respective .Rproj, with functions stored in the respective code folders and data stored in the respective data folders (which is gitignored).

---

# Question 1: Coffee Hub

## Approach

A coffee entrepreneur wants to know which origin, roast strength, and supplier to stock for a new shop, and how that lines up with what Stellenbosch students say they like in their coffee.

I approached this by building one function per plot, namely origin vs cost, roast strength, top suppliers, and a keyword-match check against the student survey wordcloud. I cleaned the data first (checked for missing values and tidied a few wide-format columns into long form), then built each plot to answer one specific part of the brief rather than trying to cram everything into a single chart.

```{r q1 setup, include=FALSE}
library(tidyverse)

list.files('Question1/code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>%
    as.list() %>% walk(~source(.))

Coffee <- silent_read(path = "Question1/data/Coffee/Coffee.csv")
```

### Data prep

First I ran unique() %>% sort() on the origin and roaster columns to spot any duplicates or inconsistencies. Roaster near-duplicates (things like curly vs straight apostrophes, or "Cafe" vs "Café") were checked but left as is because I confirmed they don't affect the top 10 supplier results since none of the split pairs have enough reviews to change the rankings. Origins needed more attention though, since inconsistent suffixes (e.g. "Chiriqui Growing Region" vs "Chiriqui Province" vs "Chiriquí") were visibly splitting what should be the same place into separate groups in the plots (which I discovered while plotting). I stripped the suffixes using gsub() and standardised the accented characters I spotted from the output. Also tidied origin_1/origin_2 and desc_1/desc_2/desc_3 from wide to long format since each is really one variable spread across multiple columns.

```{r data work, include=FALSE}

# Check for missing values/NA's
Coffee %>% summarise(across(everything(), ~sum(is.na(.))))

Coffee_long <- Coffee %>%
    gather(origin_rank, origin, origin_1, origin_2) %>%
    gather(desc_rank, desc, desc_1, desc_2, desc_3) %>%
    filter(!is.na(desc))  # two coffees are missing a third review so those rows are dropped

# roaster has some near-duplicate spellings (accents/apostrophes)
# I checked whether splits affected top 10 supplier plot results and confirmed they do not, so left as is
Coffee %>% pull(roaster) %>% unique() %>% sort()

# Chiriquí was splitting into 2 rows under different spellings, messing up the "best rated" label so I fixed here so it applies to all the origin plots
Coffee_long <- Coffee_long %>%
    mutate(origin = gsub(" Growing Region| Growing District| District| Department| County| Province| Zone| Region", "", origin)) %>%
    mutate(origin = gsub("í", "i", origin))
```

### Origin vs cost

I plotted cost against rating to see if price actually tracks quality. Bubble size shows review count, so a region's position isn't trusted equally regardless of how many reviews back it up.

Northern Sumatra is both the highest-rated origin and inexpensive and Cobán shows a well-rated coffee can be found for under $4 (high rating doesn't require a high price here).

```{r, echo=FALSE}
origin_cost_scatter(Coffee_long)
```

### Roast strength

A boxplot with individual reviews jittered on top shows both the typical rating per roast and how much it varies. This is useful since a roast could have a good median but huge spread.

Light and Medium-Light roasts consistently outscore Medium, Medium-Dark, and especially Dark roasts. 

```{r, echo=FALSE}
roast_plot(Coffee)
```

### Top suppliers

Ranked suppliers by rating directly, since that's the simplest way to answer "who should she buy from". I filtered to a review-count minimum so high ratings (with a low number of them) don't dominate. I also added a cost ceiling so the list stays realistic for a coffee shop in the Neelsie (staff and student target audience) having to compete with places like MyBrew which sell relatively cheap coffee.

Top-rated suppliers under $20/100g with at least 5 reviews each.

```{r, echo=FALSE}
supplier_plot(Coffee)
```

### Matching student preferences

The brief gave a wordcloud of words Stellenbosch students used to describe coffee they liked. I picked the words that stood out most from the wordcloud (chocolate, aroma, savory, toned, sweetly, tart) and used grepl() to check which origins' reviews mention them most frequently. I filtered to origins with at least 20 reviews so one good review review couldn't skew the result, and plotted the top 10 as a share of total reviews mentioning at least one of the keywords.

Reviews from Yemen, Tarrazu, and Boquete mention student-preferred flavour words (chocolate, aroma, savory, toned, sweetly, tart) most often.

```{r, echo=FALSE}
keyword_plot(Coffee_long)
```

---

# Question 2: Baby Names

# Approach

This question asked for the analysis of US baby naming trends (1910-2014) for a toy design agency.The agency propositioned this in order to understand what drives naming trends and how long they persist, to inform toy character naming decisions.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
list.files('Question2/code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>%
    as.list() %>% walk(~source(.))

BabyNames <- read_rds("Question2/data/US_Baby_names/Baby_Names_By_US_State.rds")
Billboard <- read_rds("Question2/data/US_Baby_names/charts.rds")
HBO_titles <- read_rds("Question2/data/US_Baby_names/HBO_titles.rds")
HBO_credits <- read_rds("Question2/data/US_Baby_names/HBO_credits.rds")
```

```{r inspect, include=FALSE}
#Have a look at each dataset to see what I am working with
glimpse(BabyNames)
glimpse(Billboard)
glimpse(HBO_titles)
glimpse(HBO_credits)

# Check for NA values
BabyNames %>% summarise(across(everything(), ~sum(is.na(.))))
```


## Naming Persistence Over Time

calc_rank_persistence() takes the top 25 names for a given year and sex, then looks up each of those names in the full national ranking three years later however far down they may have fallen. This is better than comparing two top-25 lists directly (which was my initial appraoch), because names that dropped out of the top 25 still get penalised rather than being silently excluded. persistence_plot() then plots the resulting Spearman correlation for every year from 1910 to 2011, separately for boys and girls.

```{r persistence-calc and plot}
Years <- BabyNames$Year %>% unique() %>% sort()
Years <- Years[Years <= max(Years) - 3]

PersistenceResult <- expand_grid(Year = Years, Gender = c("M", "F")) %>%
    mutate(Corr = map2_dbl(Year, Gender, ~calc_rank_persistence(BabyNames, .x, .y)))

persistence_plot(PersistenceResult)
```

- Persistence dropped and became more volatile after 1990 (for both sexes)
- Girls' names persistence started weakening earlier (mid-1950s/60s, not 1990)
- Boys' names persistence stayed stable until about 1990, then matched the same pattern
- Persistence is mostly weaker and more volatile since 1990 than in the early decades (for both sexes)

## Naming Surges and Cultural Events

The brief's tip was to look for year-on-year surges and dig into what might explain them, with Whitney in the 1980s given as the model example. So I built find_name_surges() to find the biggest single-year jumps in national popularity.

```{r find-surges}
NameSurges <- find_name_surges(BabyNames, top_n = 15)
NameSurges
```

- Several biggest male surges (Robert, John, James, Richard, Michael) all
  happen in 1946, suggesting a shared cause rather than separate ones
    - Likely the post-WWII baby boom, which raised total births that year
- Female surges (Linda, Shirley, Ashley) are more spread out, suggesting
  separate, name-specific causes
    - Shirley likely tracks child star Shirley Temple's rise in the 1930s
    - Linda and Ashley probably follow similar celebrity-driven trends,
      though not individually verified

## More Recent Rising Names

To zoom into the most recent decade of data, I plotted the names with the biggest average rise in popularity between 2005 and 2014, separately for girls and boys, using my existing plot_rising_names() function. Rather than just listing the names with a single surge year, this shows their actual year-by-year counts as lines so it's possible to see each name rose, not just that it rose.

# More Recent Rising Names
```{r rising names per gender plots, echo=FALSE}
plot_rising_names(BabyNames, gender_input = "F", from_year = 2005, to_year = 2014, top_n = 5)


plot_rising_names(BabyNames, gender_input = "M", from_year = 2005, to_year = 2014, top_n = 5)
```

- 2005 to 2014's fastest-rising names (Sophia, Harper, Liam, Mason, etc.) show       steady multi-year growth rather than a single sharp spike
- Daleyza stands out as a name with zero history before 2012, then immediate growth   (a different pattern from names simply gaining popularity over time)
  - Interestingly, a quick google search attributes this rise in the name to the Mexican-American singer (Larry Hernandez) naming his daughter in 2010

One name in this list genuinely surprised me was **Daleyza**. I'd never even heard this name before, and it doesn't follow the same pattern as most of the other surges on the list because there's basically zero history before it suddenly shows up, rather than slowly building up like a normal "rising" name. That jump from nowhere is what had me running to google and it turns out Mexican-American singer Larry Hernandez named his daughter Daleyza in 2010, which seems to line up with the name appearing out of nowhere shortly after. He also had a reality show called LarryMania - the more you know...

This prompted me to check it out:

```{r Daleyza-inspect}
Daleyza_by_state <- BabyNames %>% filter(Name == "Daleyza") %>%
    group_by(State) %>% summarise(Total = sum(Count)) %>% arrange(desc(Total))

Daleyza_by_state
```

And suprise, suprise! Daleyza's spike is heavily concentrated in California and Texas, with Arizona, Illinois, and Colorado following behind them. This matches the Larry Hernandez explanation because many Mexican-Americans live in thse states (but they are also just some of the most densely populated states in the U.S.). It is, however, important to note that 505 Daleyza's is not that many in the bigger scheme of things, but the rise in the name can still most likely be attributed to Larry Hernandez's duaghter.

## Billboard naming trends

On the questions's Whitney Houston example, I checked whether chart-topping artists' names actually correspond to baby naming surges. I built billboard_bubble_plot() to plot each year's top Billboard #1 artist first names against how many babies were given that name that year (bubble size = count). In the spirit of the Whitney example, I wanted to compare the 2000-2014 and 1980-1989 windows.

```{r billboard-bubble, echo=FALSE}
billboard_bubble_plot(BabyNames, Billboard, from_year = 2000, to_year = 2014, top_n = 5)
billboard_bubble_plot(BabyNames, Billboard, from_year = 1980, to_year = 1989, top_n = 5)
```

- Some names (Taylor in 2013, Justin in 2006, Christina/Jennifer in 2002 and 2003) show large bubbles that line up with their chart-topping years.
- Other large bubbles (Daniel, Michael, Mary) reflect names that were already extremely common regardless of any Billboard connection so from this plot it is difficult to distinguish which name surges are linked to the artists and which are genuinely common/popular names.
- Many chart-topping first names (Usher, Kanye, Adele, Bruno, Queen) appear as small dots, suggesting chart dominance alone doesn't reliably predict baby naming impact.
- The 1980s to the 2000s windows do not have very different patterns because both decades have a mix of big bubbles that belong to already common names and small bubbles of names that did not stick
    - this data doesn't provide strong evidence that naming impact from celebrities has weakened (or strengthened) over time.

I also explored using the HBO titles/credits data for a similar analysis, but the character name column contained corrupted characters that broke standard string functions, so I stuck with Billboard for this section given time constraints.


## Conclusions
- Persistence has become less predictable since the 1960s, not specifically since 1990 as suspected by the agency. 
    - Both boys' and girls' rankings show clear volatility decades before 1990, so the 1990s aren't a clean turning point, but the bigger instinct (that recent name trends are less stable than older ones) holds for the second half of the dataset.
- A practical implication for toy naming is that a name sitting in the top 25 today is a weaker guarantee of future popularity than it would have been pre-1960s so betting heavily on one trending name for a toy line carries more risk now than in earlier decades.
- Large naming surges have two driving forces, and they require different responses by the agency:
    - Population-wide events (e.g. the 1946 male-name cluster, likely tied to the post-WWII baby boom) lift already-popular names rather than creating new fashionable ones - not useful for spotting a "next big name," since nothing new is actually trending
    - Culturally-driven, name-specific spikes (Shirley Temple, Whitney Houston and also likely Daleyza/Larry Hernandez) are the ones actually worth watching. 
    - These show a single celebrity, character, or cultural moment can cause a name to gain popularity fast. 
        - This is clearly the the driver that the agency should pay attention to for toy naming purposes
- The agency should also watch out for zero-history to sudden spike names like Daleyza as this could hint to a name suddenly gaining popularity. 
- Overall, in recent years names have not been as persistent and so toy companies' should shy away from relying too much on such data.



---

# Question 3: Loans and Credit

## Approach

This question asked for an analysis of US household loan default trends using anonymised Lending Club peer-to-peer lending data. The Director of the Credit Institute in Texas wanted to understand what drives defaults and whether four beliefs of the Institute hold up against the data.

```{r setup-data, include=FALSE}
list.files('Question3/code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>%
    as.list() %>% walk(~source(.))

# Load the data
Loan_credit <- read_rds("Question3/data/Loan_Cred/loan_data.rds")

# Look at the data to assess the necessary steps for cleaning
dim(Loan_credit)

glimpse(Loan_credit)

summary(Loan_credit)

# Apply data cleaning function
Loan_credit <- clean_loans(Loan_credit)
```

The data has 1 million rows and 145 columns, so the first thing I did was reduce it to only the columns actually needed for analysis. I then built a clean_loans() function to create the outcome variables, convert text columns to numeric, and Winsorise relevant (ie. variables with crazy values as seen from the summary function) continuous variables at the 1st and 99th percentiles. Since there is no age variable in the dataset, I used credit history length derived from earliest_cr_line as a proxy since I thought that a longer history likely means an older borrower.

## Key Risk Drivers

The dataset contains many variables that could serve as default risk drivers (revolving credit utilisation (revol_util), number of public records (pub_rec), debt-to-incomeratio (dti), income verification status (verification_status), and annual income (annual_inc) are all standard credit risk indicators a lender would most likely consider. However, I focused on credit grade and loan purpose because they are the most immediately interpretable at a glance as a director looking at a bar chart of default rates by grade or by loan purpose can draw actionable conclusions without needing to understand the underlying variable in detail. 

```{r defaults-grade, echo=FALSE}
defaults_by_grade_plot(Loan_credit)
```

The plot shows a clear strong relationship between credit grade and default rate (as expected). 

```{r defaults-purpose, echo=FALSE}
defaults_by_purpose_plot(Loan_credit)
```

Car and home improvement loans carry the lowest default rates at around, while small business and wedding loans sit at the top at. Thus, lenders should apply greater scrutiny to higher-risk purpose categories when assessing borrower credibility. This plot provides evidence that loan purpose is a a good credit risk signal. 

## Looking at the Beliefs held by the Institute

The Director wishes to challenge four core beliefs held by the Institute so I tested each one using the cleaned data, building a separate plot for each belief.

The first belief was that homeowners and people employed for more than 10 years default far less on short-term loans. To test this I built a line plot showing default rates across home ownership categories, with separate lines for borrowers employed 10+ years vs under 10 years, filtering to short-term loans only since the belief is specifically about those. The data contains only two loan terms (36 and 60 months) so 36 months is used as the short-term definition.

```{r homeowner-employment, echo=FALSE}
homeowner_employment_plot(Loan_credit)
```

The plot somewhat confirms this belief since borrowers employed for 10 or more years show slightly lower default rates (lower by 1%). This difference is really small so this belief is only partly supported by the data. On the other hand, homeowners (those with over and under 10 years of employment) have higher default rates than borrowers.

The second belief was that US states differ significantly in their culture of defaulting. To test this I plotted default rates for every state ordered from lowest to highest, with Texas highlighted separately so it is easy to see where it sits relative to the rest.

```{r default_by_state, echo=FALSE}
default_by_state_plot(Loan_credit)
```

This belief is more closely supported by the data because default rates range from around 14% to about 27%/28%. This means that US states do in fact differ significantly in default culture. Texas sits close to the average when looking at the bigger picture. 

The third belief was that credit grades predict default well specifically for younger individuals. Since there is no age variable in the data I used credit history length as a proxy since shorter history likely means a younger borrower. I then plotted default rates by grade for each history group using facet_wrap, which creates one panel per group so the grade pattern can be compared side by side.

```{r grade-age, echo=FALSE}
grade_age_plot(Loan_credit)
```

The credit grading system has the same relationship as before for both younger (shorter credit history) and older (longer credit history) individuals. This relationship is that default rates increase with lower credit grades. The differences between older and younger individuals default rates' is very small. For higher grades, younger individuals have a slightly higher default rating while for higher grades it is the opposite. Thus, there is no evidence from the data to support the belief that grades work better or worse for younger borrowers specifically.

The fourth belief was that interest rates are determined by age, occupation 
and credit scores. Due to time constraints, I focused on credit grade as a proxy for credit score and plotted the median interest rate per grade. 

```{r int-rate, echo=FALSE}
interest_rate_plot(Loan_credit)
```

From this graph it is obvious that credit rating is an obvious determinant of interest rates. From the basic understanding of interest rates and credit ratings, this correlation makes sense, because a lower credit rating means there is a higher risk of default. Thus, the risk is compensated for by a higher interest rate (compared to higher credit ratings). 

I focused on credit grade as the main proxy for credit score, but managed to extend this to check whether age (with credit history length as proxy, as in the previous belief) has an independent effect.

```{r plot-int-rate-age}
interest_rate_age_plot(Loan_credit)
```

Shorter-history borrowers pay slightly higher median interest rates than longer-history borrowers at most grades. This makes logical sense, since a shorter credit history probably gives lenders less default risk information besides the credit grade itself, so a small risk premium gets added (in the form of higher interest rates). 

Occupation data does exist as emp_title, but there are over 200 000 different job titles typed in as free text, so it would need a lot of work to clean and group before it's actually useable. I didn't have time for that.

## A note on knitting

I unfortunately spent too long tweaking the Texevier YAML template for Question 3 and broke it (sorry) in the process. I get a LaTeX (pandoc) error which I will troubleshoot if I have time. The analysis and all functions run correctly in RStudio, the PDF just would not compile in time. The full write-up can be found in Question3/Question3.Rmd if needed. Update: I created a new Question3_1.rmd file that knots to pdf but I will keep my texevier file as evidence of my process.

## A note on the graphs inconsistency
Note that question 3 uses the ggplot2::theme_bw(), while all other questions use fmxdat::theme_fmx with finplot() for formatting. This is because I struggled to get the theme_fmx to work here, I am still unsure why but I think it has something to do with me breaking the Texevier template. I aimed for consistency in this question, but the graphs are now not consistent with the rest of the prac. 

---

# Question 4: Netflix

# Approach
## Data Loading and Setup

```{r q4 setup, include=FALSE}
library(fmxdat)
list.files('Question4/code/', full.names = T, recursive = T) %>%
    .[grepl('.R', .)] %>% as.list() %>% walk(~source(.))

# load raw data to see what we are working with
titles <- read_rds("Question4/data/Netflix/titles.rds")
netflix_movies <- read_csv("Question4/data/Netflix/Netflix_movies.csv")
credits <- read_rds("Question4/data/Netflix/credits.rds")

glimpse(titles)
glimpse(netflix_movies)
glimpse(credits)

# clean and prep
netflix <- clean_netflix("Question4/data/Netflix")
titles_clean <- netflix$titles
actors_clean <- netflix$actors
```

The data comes from three sources where titles.rds contains scores, genres, production countries, and runtime for all Netflix titles. credits.rds links actors and directors to titles via a shared id column. Netflix_movies.csv contains additional movie-level metadata including country, age rating, and genre categories. I used the glimpse() function as before to see what I was working with before approaching the data cleaning.

Since all figures rely on titles and credits, the cleaning focuses on those two. For titles, I filtered to movies only, removed entries with missing or zero runtimes, and used gsub() to strip the brackets and quotes from the genres and production_countries columns so they read as plain comma-separated text. For credits I simply filtered to actors only as this is what will be useful for my plots later on. 

## Genre Preferences by Country

To identify what types of movies (genre) different countries produce, I used grepl() to see genre membership per title and map_dfr() to stack results into a long dataframe. The top 10 countries by movie count were selected via count() and slice_head(), with each movie assigned to its primary production country for faceting purposes.

```{r genre-by-country, echo=FALSE}
plot_genre_by_country(titles_clean)
```

Drama and comedy seem to be the most popular genre for most countries, but there are definitely differences between countries that are worth noting. For example, Japan stands out for its unusually high share of animation, while India, Nigeria and the Philippines have a high preference for drama and romance with little prefernce for science fiction or documentation content. Thus, content strategy should not be applied at a global level, but rather tailored to the tastes of individual countries.

## Audience Ratings by Genre

To understand what content audiences rate most highly, I computed the median IMDb score per genre using the same grepl() and map_dfr() pattern as before. Note that IMDb scores are global and not Netflix-specific so they should be read as a general quality signal rather than a platform performance measure.

```{r score-by-genre, echo=FALSE}
plot_score_by_genre(titles_clean)
```

History and documentary content have the highest median scores, which suggests that focusing on such niche content could be unique selling point for a new streaming platform.

## Audience Engagement by Country

TMDb popularity scores reflect active audience engagement (that being things like searches, watchlists, and rating activity)making them a better proxy for streaming performance than critic scores alone. Given heavy right skew in the distribution (median of about 6, max  of about 1823), a log scale was applied in finplot(log.y = TRUE).I noticed this after first plotting without the log scale and the results looking not so nice.

```{r popularity-by-country, echo=FALSE}
plot_popularity_by_country(titles_clean)
```

Japanese and English-language content (GB, US) drives the highest median engagement globally, while Nigerian and Filipino content (while still relatively popular) remains more locally concentrated.

## Movie Runtime by Country

I compared runtime distributions across the top 10 production countries using a boxplot with jitter overlay, to address the brief's requirement of looking at the runtimes. Each movie is assigned to its primary (first-listed) production country for grouping purposes. This is a deliberate simplification documented here as co-productions are counted under their lead country only. The fct_reorder() function orders countries by median runtime so the plot reads cleanly from shortest to longest.

```{r runtime-by-country, echo=FALSE}
plot_runtime_by_country(titles_clean)
```

Indian cinema has the longest median runtime by a clear margin, which matches what we know about the famously long format of Bollywood films. The US has the widest spread reflecting its greatly varied runtimes. A sweetspot is revealed though this data as the other countries seemingly cluster around the 90-110 minute mark. This sweetspot should be taken into account when choosing content for the new streaming platform.

## Does Star Power Drive Audience Engagement?

When I read the question for the first time, my curiosity was definitely piqued and I immediately knew I wanted to do something with either the actors or directors in the datasets. I landed on exploring whether featuring well-known actors influences audience engagement because I saw this as a practically relevant question for a streaming platform deciding where to allocate its content budget. I defined "star power" as any movie featuring at least one of the top 50 actors ranked by total IMDb votes across their filmography. I identified them through an inner_join() between actors_clean and titles_clean on the shared id column. TMDb popularity was chosen as the outcome variable over IMDb score because it reflects active audience engagement (searches, watchlists, rating activity) rather than critical opinion. A log scale was applied, finplot(log.y = TRUE), for the same reason as earlier.

```{r star-power, echo=FALSE}
plot_star_power(titles_clean, actors_clean)
```

Star cast movies show have a visibly higher median TMDb popularity than non-star cast movies. Thus, it suggests that casting recognisable names will drive audience engagement on any new streaming platform. 

## Conclusions

- Drama and comedy are safe bets globally, but content strategy should be tailored for each region eg. animation drives huge engagement in Japan.
- History and documentary genres score highest with audiences despite low production volumes so promoting such titles might give a platform a competitive advantage.
- English-language (GB) and Japanese content gets the highest TMDb engagement scores, these languages should be considered when making global content decisions.
- Most markets cluster around a 90–110 minute runtime sweet spot so acquiring content within this window will be beneficial in keeping audiences engaged (across all regions).
- Movies featuring top IMDb-voted actors show measurably higher audience engagement, suggesting that investing in recognisable talent is a worthwhile strategy for building an initial subscriber base.

---


