---
title: "Coffee Hub Recommendations"
author: "24693197"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
list.files('code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>% as.list() %>%
    walk(~source(.))
Coffee <- silent_read(path = "data/Coffee/Coffee.csv")
```

```{r tidy data, include=FALSE}
glimpse(Coffee)

# Check for missing values/NA's
Coffee %>% summarise(across(everything(), ~sum(is.na(.))))

Coffee_long <- Coffee %>%
    gather(origin_rank, origin, origin_1, origin_2) %>%
    gather(desc_rank, desc, desc_1, desc_2, desc_3) %>%
    filter(!is.na(desc))  # two coffees are missing a third review so those rows are dropped

# roaster has some near-duplicate spellings (accents/apostrophes)
# I checked whether splits affected top 10 supplier plot results and confirmed they do not, so left as is
Coffee %>% pull(roaster) %>% unique() %>% sort()

# Chiriquí was splitting into 2 rows under different spellings, messing up the "best rated" label so I fixed here so it applies to all the origin plots
Coffee_long <- Coffee_long %>%
    mutate(origin = gsub("Growing Region|Growing District|District|Department|County|Province|Zone|Region", "", origin),
           origin = gsub("í", "i", origin),
           origin = gsub("á", "a", origin),
           origin = gsub("ú", "u", origin),
           origin = gsub("ñ", "n", origin))

dim(Coffee_long)
```

# Introduction
 - Report uses coffee review database (2,095 reviews across origins, roasters, and roast strengths)
 - Generates various plots in order to help the Entrepreneur identify the best growing region, roast strength and supplier (to match student-tastes) for the new Neelsie shop

# Origin vs. Cost Scatter
Northern Sumatra is both the highest-rated origin and inexpensive, while Cobán shows a well-rated coffee can be found for under $4. Thus, high rating doesn't require a high price in this data.

```{r origin-plot, echo=FALSE}
origin_cost_scatter(Coffee_long)
```

# Roast Strength
Light and Medium-Light roasts consistently outscore Medium, Medium-Dark, and especially Dark roasts.

```{r roast-plot, echo=FALSE}
roast_plot(Coffee)
```

# Top Suppliers
Top-rated suppliers under $20/100g with at least 5 reviews each.

```{r supplier-plot, echo=FALSE}
supplier_plot(Coffee)
```

# Matching Student Preferences
Reviews from Yemen, Tarrazu, and Boquete mention student-preferred *flavour* words (chocolate, aroma, savory, toned, sweetly, tart).

```{r keyword-plot, echo=FALSE}
keyword_plot(Coffee_long)
```

# Recommendations
- Roast Strength: Aim to prioritise Light/Medium-Light across all stock.
- Stellenbosch Students Taste Match: Kochere, Yemen, and Boquete best match what Stellenbosch students say they like.
- Top pick: Northern Sumatra, since it is top-rated and surprisingly affordable.