---
# IMPORTANT: Change settings here, but DO NOT change the spacing AT ALL.
# Remove comments and add values where applicable.
# The descriptions below should be self-explanatory

title: "Predictors of loan defaults"
#subtitle: "This will appear as Right Header"

documentclass: "elsarticle"

# --------- Thesis title (Optional - set to FALSE by default).
# You can move the details below around as you please.
Thesis_FP: FALSE
# Entry1: "An unbelievable study with a title spanning multiple lines."
# Entry2: "\\textbf{Some Guy}" # textbf for bold
# Entry3: "A thesis submitted toward the degree of Doctor of Philosophy"
# Uni_Logo: Tex/Logo.png # Place a logo in the indicated location (from your root, e.g. defaults to ~/Tex/Logo.png) and uncomment this line. Leave uncommented for no image
# Logo_width: 0.3 # If using a logo - use this to set width (size) of image
# Entry4: "Under the supervision of: \\vfill Prof. Joe Smith and Dr. Frank Smith"
# Entry5: "Stellenbosch University"
# Entry6: April 2020
# Entry7:
# Entry8:

# --------- Front Page
# Comment: ----- Follow this pattern for up to 5 authors
AddTitle: TRUE # Use FALSE when submitting to peer reviewed platform. This will remove author names.
Author1: "24693197" # First Author - note the thanks message displayed as an italic footnote of first page.
Ref1: "Stellenbosch University, South Africa" # First Author's Affiliation
Email1: ""

CorrespAuthor_1: TRUE  # If corresponding author is author 3, e.g., use CorrespAuthor_3: TRUE

# Comment out below to remove both. JEL Codes only given if keywords also given.
keywords: "Multivariate GARCH \\sep Kalman Filter \\sep Copula" # Use \\sep to separate
JELCodes: "L250 \\sep L100"

# ----- Manage headers and footers:
#BottomLFooter: $Title$
#BottomCFooter:
#TopLHeader: \leftmark # Adds section name at topleft. Remove comment to add it.
BottomRFooter: "\\footnotesize Page \\thepage" # Add a '#' before this line to remove footer.
addtoprule: TRUE
addfootrule: TRUE               # Use if footers added. Add '#' to remove line.

# --------- page margins:
margin: 2.3 # Sides
bottom: 2 # bottom
top: 2.5 # Top
HardSet_layout: TRUE # Hard-set the spacing of words in your document. This will stop LaTeX squashing text to fit on pages, e.g.
# This is done by hard-setting the spacing dimensions. Set to FALSE if you want LaTeX to optimize this for your paper.

# --------- Line numbers
linenumbers: FALSE # Used when submitting to journal

# ---------- References settings:
# You can download cls format here: https://www.zotero.org/ - simply search for your institution. You can also edit and save cls formats here: https://editor.citationstyles.org/about/
# Hit download, store it in Tex/ folder, and change reference below - easy.
bibliography: Tex/ref.bib       # Do not edit: Keep this naming convention and location.
csl: Tex/harvard-stellenbosch-university.csl # referencing format used.
# By default, the bibliography only displays the cited references. If you want to change this, you can comment out one of the following:
#nocite: '@*' # Add all items in bibliography, whether cited or not
# nocite: |  # add specific references that aren't cited
#  @grinold2000
#  @Someoneelse2010

# ---------- General:
RemovePreprintSubmittedTo: TRUE  # Removes the 'preprint submitted to...' at bottom of titlepage
Journal: "Journal of Finance"   # Journal that the paper will be submitting to, if RemovePreprintSubmittedTo is set to TRUE.
toc: FALSE                       # Add a table of contents
numbersections: TRUE             # Should sections (and thus figures and tables) be numbered?
fontsize: 11pt                  # Set fontsize
linestretch: 1.2                # Set distance between lines.
link-citations: TRUE            # This creates dynamic links to the papers in reference list.

### Adding additional latex packages:
# header-includes:
#    - \usepackage{colortbl} # Add additional packages here.

output:
  pdf_document:
    keep_tex: TRUE
    template: Tex/TexDefault.txt
    fig_width: 3.5 # Adjust default figure sizes. This can also be done in the chunks of the text.
    fig_height: 3.5
abstract: |
  Abstract to be written here. The abstract should not be too long and should provide the reader with a good understanding what you are writing about. Academic papers are not like novels where you keep the reader in suspense. To be effective in getting others to read your paper, be as open and concise about your findings here as possible. Ideally, upon reading your abstract, the reader should feel he / she must read your paper in entirety.
---

<!-- First: Set your default preferences for chunk options: -->

<!-- If you want a chunk's code to be printed, set echo = TRUE. message = FALSE stops R printing ugly package loading details in your final paper too. I also suggest setting warning = FALSE and checking for warnings in R, else you might find ugly warnings in your paper. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 5, fig.pos="H", fig.pos = 'H')
# Note: Include = FALSE implies the code is executed, but not printed in your pdf.
# warning and message = FALSE implies ugly messages and warnings are removed from your pdf.
# These should be picked up when you execute the command chunks (code sections below) in your rmd, not printed in your paper!

# Lets load in example data, and see how this can be stored and later called from your 'data' folder.
if(!require("tidyverse")) install.packages("tidyverse")
library(fmxdat)
extrafont::loadfonts(quiet = TRUE)
library(tidyverse)


# Notice that as you are working in a .Rproj file (I am assuming you are) - the relative paths of your directories start at your specified root.
# This means that when working in a .Rproj file, you never need to use getwd() - it is assumed as your base root automatically.
pacman::p_load(tidyverse, xtable, fmxdat)
list.files('code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>% as.list() %>% walk(~source(.))

# Load the data
Loan_credit <- read_rds("data/Loan_Cred/loan_data.rds")

# Look at the data to assess the necessary steps for cleaning
dim(Loan_credit)

glimpse(Loan_credit)

summary(Loan_credit)

# Apply data cleaning function
Loan_credit <- clean_loans(Loan_credit)

```


<!-- ############################## -->
<!-- # Start Writing here: -->
<!-- ############################## -->

# Introduction \label{Introduction}

This report looks at loan default trends using Lending Club data covering one million anonymised peer-to-peer loans. It looks at the beliefs the Director wishes to challenge. Namely, that home owners and people that’s been employed more than 10 years have a far lower probability of default with short term loans, that US States differ significantly when it comes to a culture of defaulting on short term loans and that Credit Grades (A - F) provide a great indicator for whether younger individuals will default or not. Also that interest rates are also clearly determined by age, occupation and credit scores.

# Data  {-}
# Step 1: Data Cleaning and Exploration

Before any analysis, the data was reduced to the columns relevant to this analysis and several cleaning steps were applied.

- Two binary outcome variables were created from loan_status. Default_Flag takes the value 1 for loans charged off or defaulted, 0 for fully paid, and NA for current loans. These are excluded from default rate calculations as their outcome is not yet known. 

- Employment length and loan term were stored as text ("6 years", "36 months") and converted to numeric for analysis.

- Credit grade was set as an ordered factor from A (lowest risk) to G (highest risk).

- Credit history length is used a s a proxy for age since there is no direct age variable in the dataset. It is derived from earliest_cr_line and it is used as a proxy since a longer history is likely linked to an older borrower.

- Continuous variables with clear outliers were Winsorized at the 1st and 99th percentiles. DTI had impossible values ranging from -1 to 999, and annual income reached as high as $9.9 million so both required capping to prevent extreme values from distorting the results.

## Key Risk Drivers

### Default Rate by Credit Grade

```{r plot-grade, echo=FALSE, fig.cap="Default rates rise monotonically from Grade A to G, confirming the grading system captures genuine credit risk."}
defaults_by_grade_plot(Loan_credit)
```
The plot shows a clear strong relationship between credit grade and default rate (as expected). 

### Default Rate by Loan Purpose

```{r plot-purpose, echo=FALSE, fig.cap="Loan purpose varies substantially in associated default risk."}
defaults_by_purpose_plot(Loan_credit)
```

Car and home improvement loans carry the lowest default rates at around, while small business and wedding loans sit at the top at. Thus, lenders should apply greater scrutiny to higher-risk purpose categories when assessing borrower credibility. This plot provides evidence that loan purpose is a a good credit risk signal. 

## Evaluating the beliefs 

### Belief: Homeowners and Long-Tenured Employees Default Less

```{r plot-heuristic, echo=FALSE, fig.cap="The director's heuristic receives only partial support."}
homeowner_employment_plot(Loan_credit)
```

The above plot somewhat confirms this belief since borrowers employed for 10 or more years show slightly lower default rates (lower by $1%$). This difference is really small so this belief is only partly supported by the data. On the other hand, homeowners (those with over and under 10 years of employment) have higher default rates than borrowers. 

### Belief: US States Differ Significantly in Default Culture

```{r plot-state, echo=FALSE, fig.cap="Default rates vary across states but Texas sits close to the national average.", fig.width=12, fig.height=5}
default_by_state_plot(Loan_credit)
```

This belief is more closely supported by the data because default rates range from around $14%$ to about $27%/28%$. This means that US states do in fact differ significantly in default culture. Texas (labeled TX on the x-axis) sits slightly above the the average when looking at the bigger picture. 

### Belief: Credit Grades (A - F) Predict Whether Younger Individuals Will Default Well

```{r plot-grade-age, echo=FALSE}
grade_age_plot(Loan_credit)
```

The credit grading system has the same relationship as before for both younger (shorter credit history) and older (longer credit history) individuals. This relationship is that default rates increase with lower credit grades. The differences between older and younger individuals default rates' is very small. For higher grades, younger individuals have a slightly higher default rating while for higher grades it is the opposite. Thus, there is no evidence from the data to support the belief that grades work better or worse for younger borrowers specifically.

### Belief: Interest Rates Are Determined by Age, Occupation, and Credit Scores

```{r plot-int-rate, echo=FALSE}
interest_rate_plot(Loan_credit)
```

From this graph it is obvious that credit rating is an obvious determinant of interest rates. From the basic understanding of interest rates and credit ratings, this correlation makes sense, because a lower credit rating means there is a higher risk of default. Thus, the risk is compensated for by a higher interest rate (compared to higher credit ratings). 

# Conclusion

Credit grade is one of the strongest predictors of default since there is a clear relationship (accross many of the plots) between grade and default rate. Loan purpose could also act as an alternative risk signal since car and home improvement loans have the lowest default rates while small business and wedding loans have the highest.

The evidence on the four beliefs of the Institute is relatively mixed. The homeownership and length of employment belief is only partially supported because borrowers with 10 or more years of employment default slightly less, but the difference is small. The belief that US states differ in default culture is supported, with default rates ranging from around 14% to 27% across states and Texas sits close to the national average. Credit grades predict default consistently for both younger and older borrowers with very small differences between the two groups. There is no evidence that grades work better or worse for younger individuals specifically. Credit grade is clearly the most important determinant of interest rates, with age (proxied by credit history length) showing only a small additional effect on top of grade. Occupation couldn't be assessed since it exists only as messy free text that wasn't feasible to clean in time.

<!-- Make title of bibliography here: -->
<!-- \newpage -->

\newpage

# References {-}

<div id="refs"></div>


# Appendix {-}

## Appendix A {-}

Some appendix information here

## Appendix B {-}

